Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-08-20 demystify.tex
updates

    %       \item
    %       \item
    %       \item (\autoref{sec:msrequirement}), and CS PhD takes abou (\autoref{sec:time}). (\autoref{sec:non-us-differences}).
    %       \item
    %       \item
    %       \item
    %       %\item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
    %     \end{itemize}

@@

\subsection{How much do \emph{I} cost?}\label{sec:ra-cost}
\sectioninfo{Your entire PhD program costs about \$400K in total, but you \emph{do not} pay for it.}

@@

\section{Why did I get rejected?}\label{sec:why-rejected}

@@

\section{Increasing my admission chance}\label{sec:improve-your-chance}

@@

\section{How long do you evaluate an application?}\label{sec:ievaluate}
\sectioninfo{It takes about 10 minutes to review an application.}

@@

\subsection{How Decisions are Made}\label{sec:how-decisions}

@@

\section{Member of an underrepresented or LGBTQ+ groups}\label{sec:urm}
\sectioninfo{Being a minority can significantly help your application.}

@@

\section{PhD in the US vs. Other Countries}\label{sec:non-us-differences}
\sectioninfo{Compared to other countries, CS PhD in the US does not require an MS degree but has a longer PhD study time.}

@@

\section{How do I call or address a professor?}\label{sec:address}
\sectioninfo{What should you call your profs.? Many possibilities (e.g., prof., Dr., and even their firstnames), but not Mr. or Mrs.}

@@

\appendix

@@

\chapter{Domestic Students}\label{sec:domestic-students}

@@

\chapter{Research Opportunities}\label{sec:research-opportunities}

@@

\chapter{Glossary}\label{sec:glossary}

  \item[Adcom] Admission Committee. The group of faculty members who review applications and make admission decisions.
  \item [Adcom members] Faculty members who are part of the admission committee. These are the people who review applications and make admission decisions.
  \item[Adcom chair] The faculty member who leads the admission committee. This person is often not involved in individual admission decisions but oversees the process (e.g., resolve disputes, ensure fairness).
  \item[Advisor/Supervisor]: A faculty member who guides and mentors a PhD student throughout their research. This person typically plays a crucial role in a student's academic journey. In the US, the term ``adviser'' is more commonly used than ``supervisor''.
  \item[Cohort] A group of students who start a program at the same time and take classes together. You should get to know your cohort as they will be your colleagues and friends during your PhD.
  \item  [Deferred Admission] An option allowing admitted students to postpone their start date, typically by one year.
  \item [Diversity Statement] A document sometimes required in applications where applicants describe their contributions to diversity and inclusion, or how their background adds to the diversity of the academic community. This document is often required for faculty positions, but some universities also require it for graduate admissions.
  \item[Domestic Student] A student who is typically a US citizen or permanent resident (``greencard holder'').
        But sometimes it can also refer to students who did their undergrad (or MS) at a US university.
  \item[Field or Area] A specific research area, e.g., Machine Learning, Computer Vision, Software Engineering, etc.
  \item [In-state vs. Out-of-state] In-state tuition is the tuition rate for students who are residents of the state where the university is located. Out-of-state tuition is the tuition rate for students who are not residents of that state. In-state tuition is typically much lower than out-of-state tuition. PhD students typically do not have to pay tuition as it is covered by their funding, but this might be important for MS students.
  \item[LoR] Letter of Recommendation. A letter written by a professor or supervisor that assesses your qualifications and potential for graduate study (\autoref{sec:lor}).
  \item[SOP] Statement of Purpose. A document written by yourself to explains your research interests, background, and reasons for applying to a PhD program (\autoref{sec:sop}).

  \item [Lab] Research Lab. A group of researchers (profs. and students) working on a common research area, e.g., a Software Engineering Lab.  Typically a lab is led by a single professor, but sometimes it can be co-led by multiple professors. For example, my lab is DynaROARS, which focuses on my own research area (software verification), but I am also part of the more general SWE Lab at Mason with other faculty in Software Engineering.

  \item[MS] Master of Science. A graduate degree that typically requires 2 years of study.
  \item [PI]  Principal Investigator. A lead researcher on a grant or research project, e.g., a PI on an NSF grant. Prospective students often use this term to refer to a professor they are interested in working with.  While this term is used frequently in online forums among students, we do not use it this way in academia, e.g., ``who is your PI?'' might only make sense to fellow Reddit applicants, but might not be understood by others, ``who is your adviser?'' is what you want to use.
    \item[STEM] Science, Technology, Engineering, and Mathematics. A term used to group together these academic disciplines. CS is considered a STEM field.
  \item[URM] Underrepresented Minority. Refers to groups that are underrepresented in higher education, e.g., Black, Hispanic, and Native American students.
  \item[NSF] National Science Foundation. A US government agency that supports research and education in science and engineering. NSF is a major source of funding for RAs in CS research. There are other agencies that support RAs but they might require US citizenship or permanent residency.
        \item [R1 vs. R2 Universities] Research 1 (R1) universities are universities with the highest level of research activity. Research 2 (R2) universities have a high level of research activity but not as high as R1 universities.  R1 universities are typically more prestigious and have more funding and resources for research.  R2 universities are also good but might have fewer resources and funding.
  \item [Rolling Admission] A process where applications are reviewed as they are received, and decisions are made throughout the admission cycle.  This is in contrast to a process where all applications are reviewed after the deadline and decisions are made at the same time. Rolling admission is more common for MS programs and less common for PhD programs (e.g., at Mason, MS is rolling and PhD is not).
  \item [Top-tier conferences and journals] These are the most prestigious venues for publishing research in CS.  For example, in Software Engineering, top-tier conferences include ICSE, FSE, and ASE and top-tier journals include TSE and TOSEM.  Publishing in these venues is highly competitive and prestigious, and can significantly improve your chances of getting admitted to a good PhD program.

>>>>

    %       \item
    %       \item
    %       \item (\autoref{sec:msrequirement}), and CS PhD takes abou (\autoref{sec:time}). Pu(\autoref{sec:non-us-differences}).
    %       \item
    %       \item
    %       \item
    %       %\item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
    %     \end{itemize}

@@

\subsection{How much do you cost?}\label{sec:ra-cost}
\sectioninfo{Your entire PhD program costs about \$400K in total, but you \emph{do not} pay for it.}

@@

\section{Why did you get rejected?}\label{sec:why-rejected}

@@

\section{Increasing your admission chance}\label{sec:improve-your-chance}

@@

\section{How long does it take to evaluate an application?}\label{sec:ievaluate}
\sectioninfo{It takes me about 10 minutes to review an application.}

@@

\section{How are decisions made?}\label{sec:how-decisions}
\sectioninfo{Even if all adcom reviewers recommend acceptance, the application can still be rejected. Vice versa, if all reviewers think the application is weak, the student might still be admitted.}

@@

\section{Member of an underrepresented or LGBTQ+ groups}\label{sec:urm}
\sectioninfo{Being a minority can make you stand out and help your application.}

@@

\section{PhD in the US vs. Other Countries}\label{sec:non-us-differences}
\sectioninfo{Among several differences, CS PhD in the US does not require an MS degree but has a longer PhD study time.}

@@

\section{How to call or address a professor?}\label{sec:address}
\sectioninfo{What should you call your profs.? Many possibilities (e.g., prof., Dr., and even their firstnames), but not Mr. or Mrs.}

@@

\appendix
\chapter{Glossary}\label{sec:glossary}

  \item[Adcom] Admission Committee. The group of faculty members who review applications and make admission decisions.
  \item [Adcom members] Faculty members who are part of the admission committee. These are the people who review applications and make admission decisions.
  \item[Adcom chair] The faculty member who leads the admission committee. This person is often not involved in individual admission decisions but oversees the process (e.g., resolve disputes, ensure fairness).
  \item[Advisor/Supervisor]: A faculty member who guides and mentors a PhD student throughout their research. This person typically plays a crucial role in a student's academic journey. In the US, the term ``adviser'' is more commonly used than ``supervisor''.
  \item[Cohort] A group of students who start a program at the same time and take classes together. You should get to know your cohort as they will be your colleagues and friends during your PhD.
  \item  [Deferred Admission] An option allowing admitted students to postpone their start date, typically by one year.
  \item [Diversity Statement] A document sometimes required in applications where applicants describe their contributions to diversity and inclusion, or how their background adds to the diversity of the academic community. This document is often required for faculty positions, but some universities also require it for graduate admissions.
  \item[Domestic Student] A student who is typically a US citizen or permanent resident (``greencard holder'').
        But sometimes it can also refer to students who did their undergrad (or MS) at a US university.
  \item[Field or Area] A specific research area, e.g., Machine Learning, Computer Vision, Software Engineering, etc.
  \item [In-state vs. Out-of-state] In-state tuition is the tuition rate for students who are residents of the state where the university is located. Out-of-state tuition is the tuition rate for students who are not residents of that state. In-state tuition is typically much lower than out-of-state tuition. PhD students typically do not have to pay tuition as it is covered by their funding, but this might be important for MS students.
  \item [Lab] Research Lab. A group of researchers (profs. and students) working on a common research area, e.g., a Software Engineering Lab.  Typically a lab is led by a single professor, but sometimes it can be co-led by multiple professors. For example, my lab is DynaROARS, which focuses on my own research area (software verification), but I am also part of the more general SWE Lab at Mason with other faculty in Software Engineering.

  \item[LoR] Letter of Recommendation. A letter written by a professor or supervisor that assesses your qualifications and potential for graduate study (\autoref{sec:lor}).
  \item[MS] Master of Science. A graduate degree that typically requires 2 years of study.
  \item[Open House] An event where admitted students visit the university to meet faculty, students, and staff, and learn more about the program. This is a great opportunity to get a feel for the university and the department and decide if it is a good fit for you.

  \item[SOP] Statement of Purpose. A document written by yourself to explains your research interests, background, and reasons for applying to a PhD program (\autoref{sec:sop}).

  \item [PI]  Principal Investigator. A lead researcher on a grant or research project, e.g., a PI on an NSF grant. Prospective students often use this term to refer to a professor they are interested in working with.  While this term is used frequently in online forums among students, we do not use it this way in academia, e.g., ``who is your PI?'' might only make sense to fellow Reddit applicants, but might not be understood by others, ``who is your adviser?'' is what you want to use.
    \item[STEM] Science, Technology, Engineering, and Mathematics. A term used to group together these academic disciplines. CS is considered a STEM field.
  \item[URM] Underrepresented Minority. Refers to groups that are underrepresented in higher education, e.g., Black, Hispanic, and Native American students.
  \item[NSF] National Science Foundation. A US government agency that supports research and education in science and engineering. NSF is a major source of funding for RAs in CS research. There are other agencies that support RAs but they might require US citizenship or permanent residency.
        \item [R1 vs. R2 Universities] Research 1 (R1) universities are universities with the highest level of research activity. Research 2 (R2) universities have a high level of research activity but not as high as R1 universities.  R1 universities are typically more prestigious and have more funding and resources for research.  R2 universities are also good but might have fewer resources and funding.
  \item [Rolling Admission] A process where applications are reviewed as they are received, and decisions are made throughout the admission cycle.  This is in contrast to a process where all applications are reviewed after the deadline and decisions are made at the same time. Rolling admission is more common for MS programs and less common for PhD programs (e.g., at Mason, MS is rolling and PhD is not).
  \item [Top-tier conferences and journals] These are the most prestigious venues for publishing research in CS.  For example, in Software Engineering, top-tier conferences include ICSE, FSE, and ASE and top-tier journals include TSE and TOSEM.  Publishing in these venues is highly competitive and prestigious, and can significantly improve your chances of getting admitted to a good PhD program.

@@

\chapter{Domestic Students}\label{sec:domestic-students}
\chapterinfo{Specific benefits and opportunities for domestic students applying to CS PhD programs.}

@@

\chapter{Research Opportunities}\label{sec:research-opportunities}
\chapterinfo{How to get research experience as an undergrad.}

@@
